A Data-Driven Framework for Neural Field Modeling

### 2010-11-20 ltx/first_order_framework/LaTeX_Manuscript.tex
added research highlights

	\begin{frontmatter}
% Title must be 150 characters or less
\title{A Data-Driven Framework for Neural Field Modelling}

@@

\begin{keyword}
	\parham{keyword1 \sep keyword2}
%% keywords here, in the form: keyword \sep keyword

@@

Perhaps the most accepted estimation framework for neural mass models is dynamical causal modelling (DCM) \cite{David2003,David2006}, which has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations~\cite{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

Modelling the neural dynamics within this framework has a distinct advantage over the more standard multivariate auto-regressive (MVAR) models: the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity.

In this paper, we demonstrate for the first time how an intracortical connectivity kernel can be inferred from data, based on a variant of the WC neural field model~\cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and electrophysiological data. The paper proceeds by first describing the continuum neural field equations that are used as the cortical model. Then a finite-dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. The next section establishes conditions, using spatial frequency analysis, for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in the following section. The results for the spatial frequency analysis and parameter estimation are then presented. Finally, the implications and limitations of this framework are discussed along with planned future developments.

@@

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modelling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Fig.~\ref{fig:Figure1}. To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

@@

where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (and states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. \parham{Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives}

@@

\subsection{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section, we describe the procedure for estimating the states, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic dynamics, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. Once initialized, at each iteration the sequence of estimated state vectors is used to update the parameter set. The resulting parameters are then used when estimating the new state vector sequence for the next iteration. The procedure stops when the \parham{parameter estimates} converge. The algorithm is initialized using a bounded random state vector sequence, guaranteeing that the initial estimated parameter set forms a stable kernel.

@@

Previous work by Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter (UKF) for state estimation of discrete (in space and time) nonlinear spatiotemporal systems. Within their framework, each discrete spatial location formed an element of the state vector to be estimated with the UKF. A limitation of this framework was that an increase in spatial resolution and/or the physical size of the neural field resulted in an increased state dimension. This, in turn, leads to increases in uncertainty in the state estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modelling assumption.

@@

\subsection{Patient-Specific Modelling}

@@

The development of the estimation framework also has implications for theoretical neuroscience, where the ability to estimate subject-specific properties will allow testing of hypotheses that have been generated in computational modelling studies. This has the potential to validate neural field models and, in doing so, provide new insights that may lead to a greater understanding of mesoscopic and macroscopic neurodynamics. For example, using the proposed model-based framework, we can reconstruct a neural field from data and estimate parameters to check if they correspond to theoretically derived values where certain solutions may exist. In addition, the parameters and the state-space can be explored during cognitive tasks and states of arousal such as sleep and anaesthesia.

>>>>

	\begin{frontmatter}
% Title must be 150 characters or less
\title{A Data-Driven Framework for Neural Field Modeling}

@@

\begin{keyword}
	Neural Field Model \sep Nonlinear Estimation \sep Intracortical Connectivity \sep Nonlinear Dynamics
%% keywords here, in the form: keyword \sep keyword

@@

Perhaps the most accepted estimation framework for neural mass models is dynamical causal modeling (DCM) \cite{David2003,David2006}, which has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations~\cite{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

Modeling the neural dynamics within this framework has a distinct advantage over the more standard multivariate auto-regressive (MVAR) models: the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity.

In this paper, we demonstrate for the first time how an intracortical connectivity kernel can be inferred from data, based on a variant of the WC neural field model~\cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modeling and electrophysiological data. The paper proceeds by first describing the continuum neural field equations that are used as the cortical model. Then a finite-dimensional neural field model is derived. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. The next section establishes conditions, using spatial frequency analysis, for both sensor and basis function spacing and width, such that the dominant dynamics of the neural field can be represented by the reduced model. The state and parameter estimation procedure is described in the following section. The results for the spatial frequency analysis and parameter estimation are then presented. Finally, the implications and limitations of this framework are discussed along with planned future developments.

@@

\subsection{Integro-Difference Equation Neural Field Model}
The combination of modeling techniques in this paper leads to a large amount of notation, so a reference of the symbols used is provided in Table~\ref{tab:Notation}. The model relates the average number of action potentials $g(\mathbf{r},t)$ arriving at time $t$ and position $\mathbf{r}$ to the local post-synaptic membrane voltage $v(\mathbf{r},t)$. The post-synaptic potentials generated at a neuronal population at location $\mathbf{r}$ by action potentials arriving from all other connected populations is described by 
\begin{equation}
	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right) \, dt'}. 
\end{equation}
The post-synaptic response kernel $h(t)$ is described by 
\begin{equation}
	\label{SynapticRespKernel} h(t) = \eta(t)\exp{\left(-\zeta t\right)}, 
\end{equation}
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function. Non-local interactions between cortical populations at positions $\mathbf{r}$ and $\mathbf{r}'$ are described by 
\begin{equation}
	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\, d\mathbf{r}'}, 
\end{equation}
where $f(\cdot)$ is the firing rate function, $w(\cdot)$ is the spatial connectivity kernel and $\Omega$ is the spatial domain representing a cortical sheet or surface. The connectivity kernel is typically a ``Mexican hat'' or ``wizard hat'' function, which describes local excitation and surround inhibition. An example of the Mexican hat connectivity kernel used in this paper is shown in Fig.~\ref{fig:Figure1}. To demonstrate flexibility in the estimation algorithm and to model larger cortical regions, a third component of the kernel is introduced describing weak longer-range excitation. The exact shape of this kernel is assumed to vary across subjects, and hence needs to be inferred from subject-specific data.

@@

where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (and states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives

@@

\subsection{State and Parameter Estimation}\label{StateAndParameterEstimationSection} In this section, we describe the procedure for estimating the states, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic dynamics, $\xi$. The estimation process is a two part iterative algorithm, consisting of a state estimation step followed by a parameter estimation step. Once initialized, at each iteration the sequence of estimated state vectors is used to update the parameter set. The resulting parameters are then used when estimating the new state vector sequence for the next iteration. The procedure stops when the parameter estimates converge. The algorithm is initialized using a bounded random state vector sequence, guaranteeing that the initial estimated parameter set forms a stable kernel.

@@

Previous work by Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter (UKF) for state estimation of discrete (in space and time) nonlinear spatiotemporal systems. Within their framework, each discrete spatial location formed an element of the state vector to be estimated with the UKF. A limitation of this framework was that an increase in spatial resolution and/or the physical size of the neural field resulted in an increased state dimension. This, in turn, leads to increases in uncertainty in the state estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modeling assumption.

@@

\subsection{Patient-Specific Modeling}

@@

The development of the estimation framework also has implications for theoretical neuroscience, where the ability to estimate subject-specific properties will allow testing of hypotheses that have been generated in computational modeling studies. This has the potential to validate neural field models and, in doing so, provide new insights that may lead to a greater understanding of mesoscopic and macroscopic neurodynamics. For example, using the proposed model-based framework, we can reconstruct a neural field from data and estimate parameters to check if they correspond to theoretically derived values where certain solutions may exist. In addition, the parameters and the state-space can be explored during cognitive tasks and states of arousal such as sleep and anaesthesia.

### 2010-11-20 ltx/first_order_framework/SubmissionLetter.tex
added research highlights

\begin{document}
\vspace{-1cm}
\begin{letter}{Editor/Reviewer \\ PLoS Computational Biology}

@@

I am writing to seek publication of our research article titled `A Data-Driven Framework for Neural Field Modelling' in PLoS Computational Biology.

This paper brings several new techniques to the field of computational biology, where as the title suggests, we have developed a data-driven framework for estimating the parameters of a neural field model from electrophysiological data. Our aim in this work is to model the intracortical connectivity and synaptic dynamics of an individual patient or experimental subject. These subject-specific neural field models will provide new avenues for the application of engineering techniques to neural systems, particularly to epilepsy and other neurological disorders, and to brain-computer interfaces.

@@

Due to the novel aspects of the model reduction and estimation procedure, we have chosen to precede the Results section with the Methods section.

>>>>

\begin{document}
\vspace{-1cm}
\begin{letter}{Editor/Reviewer \\ Neuroimage}

@@

I am writing to seek publication of our research article titled `A Data-Driven Framework for Neural Field Modeling' in Neuroimage.

This paper brings several new techniques to the fields of neuroscience and neuroimaging, where as the title suggests, we have developed a data-driven framework for estimating the parameters of a neural field model from electrophysiological data. Our aim in this work is to model the intracortical connectivity and synaptic dynamics of an individual patient or experimental subject. These subject-specific neural field models will provide new avenues for the application of engineering techniques to neural systems, particularly to epilepsy and other neurological disorders, and to brain-computer interfaces.

@@
